---
title: "Sars-CoV-2 Spike Mutations"
author: "Elliot Hershberg"
date: "2/21/2021"
output: html_document
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(tidyverse)
library(JBrowseR)
library(r3dmol)
knitr::opts_chunk$set(echo = FALSE)
```

```{r context="data", message=FALSE}
mutation_data <- read_tsv("spike-mutations.tsv")
```

## Introduction

This is an interactive Shiny document intended to provide useful ways to reason about data concerning mutations
to the Sars-CoV-2 spike glycoprotein.
Understanding the functional significance of these new mutations is crucial for genomic epidemiology
and vaccine development.

## Spike Mutation Data

Below is a table of mutations to the spike protein curated by [Dr. Emma Hodcroft](https://twitter.com/firefoxx66):

```{r}
DT::dataTableOutput("mutationsTable")
```

```{r, context="server"}
output$mutationsTable <- DT::renderDataTable(
  {
    mutation_data
  },
  selection = "single",
  options = list(
    columnDefs = list(list(className = "dt-center", targets = "_all"))
  )
)

# set up event listener on table selection
table_selection <- reactive({
  mutation_data[input$mutationsTable_rows_selected, ]
})
```

## Browsing Genomic Context

This is an interactive JBrowse 2 genome browser embedded using the [JBrowseR](https://elliothershberg.github.io/JBrowseR/) package.
Clicking on one of the mutations in the table above will navigate to that location in the browser.

```{r}
tags$div(
  JBrowseROutput("browserOutput"),
  style = "overflow: auto;"
)
```

```{r, context="server"}
# create the necessary JB2 assembly configuration
assembly <- assembly(
  "https://jbrowse.org/genomes/sars-cov2/fasta/sars-cov2.fa.gz",
  bgzip = TRUE
)

# create configuration for a JB2 GFF FeatureTrack
annotations_track <- track_feature(
  "https://jbrowse.org/genomes/sars-cov2/sars-cov2-annotations.sorted.gff.gz",
  assembly
)

mutations_track <- track_feature(
  "https://jbrowse.org/genomes/sars-cov2/data/sars-cov2-spike-mutations.gff3.gz",
  assembly
)

# create the tracks array to pass to browser
tracks <- tracks(
  annotations_track,
  mutations_track
)

# set up the default session for the browser
default_session <- default_session(
  assembly,
  c(annotations_track)
)

theme <- theme("#5da8a3", "#333")

location <- reactive({
  print(table_selection())
  if (is.na(table_selection()$chrom_start[1])) {
    "NC_045512.2:1..100"
  } else {
    str_glue(
      "NC_045512.2:",
      "{table_selection()$chrom_start}",
      "..",
      "{table_selection()$chrom_end}"
    )
  }
})


# link the UI with the browser widget
output$browserOutput <- renderJBrowseR(
  JBrowseR(
    "View",
    assembly = assembly,
    tracks = tracks,
    location = location(),
    defaultSession = default_session,
    theme = theme
  )
)
```

## Exploring Protein Structure

Here is an embedded interactive [3Dmol.js](https://3dmol.csb.pitt.edu/) protein model of the (closed state)
Sars-CoV-2 spike glycoprotein from [PDB](https://www.rcsb.org/structure/6vxx).
It was created using the [r3dmol](https://swsoyee.github.io/r3dmol/) package.
The model can be dragged, and zoomed.
When a row is selected in the table of mutations above, the stick model for that residue will be activated.
Give it a try!

```{r}
r3dmolOutput("spikeProtein")
```

```{r, context="server"}
output$spikeProtein <- renderR3dmol(
  spike_model <- r3dmol() %>%
    m_add_model(data = m_fetch_pdb("6vxx"), format = "pdb") %>%
    m_set_style(style = m_style_cartoon()) %>%
    m_add_style(
      style = c(
        m_style_stick(),
        m_style_sphere(scale = 0.3)
      ),
      sel = m_sel(
        resi = table_selection()$mutation_pos
      )
    ) %>%
    m_zoom_to()
)
```

## Conclusion

This document provides a demonstration of how Shiny and R Markdown can be used together
with dynamic visualization libraries to communicate scientific results interactively.
Importantly, this type of interactive data visualization doesn't require hundreds of
lines of JavaScript. 
Instead, it can be done with <100 lines of R code!
Check out the source code for this document 
[here](https://github.com/elliothershberg/shiny-sars-cov-2/blob/main/shiny_doc/sars-cov-2-spike-mutations.Rmd)

In order to be more widely used by publications, more infrastructure efforts such as eLife's 
[reproducible article](https://elifesciences.org/labs/ad58f08d/introducing-elife-s-first-computationally-reproducible-article) will need to take place in biomedical journals.
Perhaps in the future it will be commonplace for genomics articles to contain interactive
genome browsers displaying the direct read support for their findings!

